Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap

We can clearly see that this is a refinement type system in the sence of
\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)

@@

Another interesting use of the trace monad to build a sophisticated \isotopessa
model is to build a model of TSO-style weak memory based on \citet{sparky}. For
simplicity, we will consider an infinite set of named locations \(x, y, z \in
\ms{Loc}\) which are subject to concurrent modification by all threads via TSO
reads, writes, and fences. We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
  \to \mc{A}\) quotiented over the removal of finitely many \(\delta\). A pomset
  is \textbf{finite} if its carrier set is. We will write unordered pomsets
  using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets
  using list notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
predecessors. They also form a monoid under parallel composition, defining
\(\alpha || \beta = [\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given
the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
Note in both cases the monoidal unit is \(\{\delta\}\), since we only consider
nonempty pomsets but allow the removal of finitely many ticks. Given any
partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in
N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the
function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega
\to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on
finite pomsets.

We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}}
= \mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is taken as a stream action on finite
program order pomsets. This yields an \isotopessa model with support for
concurrent read and write operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1},
\ms{Word})\), \(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\(
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
\),
\(
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\).
% The semantics of \(\ms{read}\) in particular give a hint as to how this model
% works: rather than tracking the state of the heap, we simply \textit{emit} a
% pomset of the events that would have be generated by a given execution (in the
% case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
% the case of a write, the single write event \(x := v\) for a write of \(v\)),
% and later post-filter to obtain a set of valid executions. By doing so, our
% semantics remains compositional, allowing us to reason about each program
% fragment individually, while still allowing other program fragments to perform
% concurrent operations affecting our potential executions. 
% On that note, 
We can define the \textit{parallel execution} of two morphisms as follows:
\begin{equation}
  \begin{aligned}
  f_0 || f_1 = \lambda (a_0, a_1). 
  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms to obtain a
monoidal (rather than preomonoidal) category, but unfortunately, sliding still
fails: \((W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})\) emits the
pomset \([x := a, y := b]\), while \(\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}}
|| \ms{id})\) emits the pomset \([y := b, x := a]\). 
% If we did want this behaviour, we'd need to have a significantly more
% sophisticated model of composition which effectively takes nontermination into
% account, which we will leave to future work.

To graduate from sequential consistency to a genuine, if maximally simple, weak
memory model, we introduce \textit{load buffering} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. On the other hand, read events will first attempt
to read from the buffer, and, if there is no corresponding write in the buffer,
will read an arbitrary value, in both cases pushing an event to the global
pomset. At any point, we may choose to \textit{flush} some of the writes in
buffer. We introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of
\textit{buffer write actions}, where an action \(\bufloc{x} := v\) by a thread
denotes adding a write \(x := v\) to the thread's write buffer. We may then
define the set of TSO actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup
\mc{A}_b\); a pomset over this set is called a \textbf{TSO pomset}. A
buffer will be defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\),
which we will interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by
index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
where \(\Sigma\) is taken as a the stream action of finite TSO pomsets on TSO
pomsets. We can view \(\ms{PO}\) as a submonad of this monad.

>>>>

We can clearly see that this is a refinement type system in the sense of
\citet{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)

@@

Another interesting use of the trace monad to build a sophisticated \isotopessa
model is to build a model of TSO-style weak memory based on \citet{sparky}. For
simplicity, we will consider an infinite set of named locations \(x, y, z \in
\ms{Loc}\) which are subject to concurrent modification by all threads via TSO
reads, writes, and fences. We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
  \to \mc{A}\) quotiented over the removal of finitely many \(\delta\). A pomset
  is \textbf{finite} if its carrier set is. We will write unordered pomsets
  using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets
  using list notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
\(\ms{trim}(R)\) removes all elements of \(R\) with infinitely many
predecessors. They also form a monoid under parallel composition, defining
\(\alpha || \beta = [\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given
the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
Note in both cases the monoidal unit is \(\{\delta\}\), since we only consider
nonempty pomsets but allow the removal of finitely many ticks. Given any
partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in
N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the
function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega
\to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on
finite pomsets.

We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}}
= \mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is taken as a stream action on finite
program order pomsets. This yields an \isotopessa model with support for
concurrent read and write operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1},
\ms{Word})\), \(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\(
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
\),
\(
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\).
% The semantics of \(\ms{read}\) in particular give a hint as to how this model
% works: rather than tracking the state of the heap, we simply \textit{emit} a
% pomset of the events that would have be generated by a given execution (in the
% case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
% the case of a write, the single write event \(x := v\) for a write of \(v\)),
% and later post-filter to obtain a set of valid executions. By doing so, our
% semantics remains compositional, allowing us to reason about each program
% fragment individually, while still allowing other program fragments to perform
% concurrent operations affecting our potential executions. 
% On that note, 
We can define the \textit{parallel execution} of two morphisms as follows:
\begin{equation}
  \begin{aligned}
  f_0 || f_1 = \lambda (a_0, a_1). 
  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms to obtain a
monoidal (rather than preomonoidal) category, but unfortunately, sliding still
fails: \((W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})\) emits the
pomset \([x := a, y := b]\), while \((\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}}
|| \ms{id})\) emits the pomset \([y := b, x := a]\). 
% If we did want this behaviour, we'd need to have a significantly more
% sophisticated model of composition which effectively takes nontermination into
% account, which we will leave to future work.

To graduate from sequential consistency to a genuine, if maximally simple, weak
memory model, we introduce \textit{load buffering} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. On the other hand, read events will first attempt
to read from the buffer, and, if there is no corresponding write in the buffer,
will read an arbitrary value, in both cases pushing an event to the global
pomset. At any point, we may choose to \textit{flush} some of the buffered
writes. We introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of
\textit{buffer write actions}, where an action \(\bufloc{x} := v\) by a thread
denotes adding a write \(x := v\) to the thread's write buffer. We may then
define the set of TSO actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup
\mc{A}_b\); a pomset over this set is called a \textbf{TSO pomset}. A buffer
will be defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which
we will interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index
(with the empty list corresponding to \(\{\delta\}\)). In particular, we define
the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where
\(\Sigma\) is taken as a the stream action of finite TSO pomsets on TSO pomsets.
We can view \(\ms{PO}\) as a submonad of this monad.
